---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# cfid: An R Package for Identification of Counterfactual Queries in Causal Models

<!-- Badges -->
[![codecov](https://codecov.io/gh/santikka/cfid/branch/main/graph/badge.svg?token=13KFY7ULZ4)](https://codecov.io/gh/santikka/cfid)
[![R-CMD-check](https://github.com/santikka/cfid/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/cfid/actions)


```{r, echo = FALSE}
library(cfid)
```

## Overview

Facilitates the identification of counterfactual queries in structural causal 
models via the ID* and IDC* algorithms by Shpitser, I. and Pearl, J. (2007, 2008) 
https://arxiv.org/abs/1206.5294,
https://jmlr.org/papers/v9/shpitser08a.html. Provides a simple interface for 
defining causal graphs and counterfactual conjunctions. Construction of parallel
worlds graphs and counterfactual graphs is done automatically based on the 
counterfactual query and the causal graph.

## Installation

Install the latest development version by using the devtools package:
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("santikka/cfid")
```

## Graphs

Directed acyclic graphs (DAG) can be defined using the function `dag` in a syntax
similar to the [`dagitty`](https://cran.r-project.org/package=dagitty) package.
This function accepts
edges of the form `X -> Y`, `X <- Y`, and `X <-> Y`, where the last variant is 
a shorthand for a latent confounder affecting both `X` and `Y` (a so-called
bidirected edge). Subgraphs can be define using curly braces `{...}`. Edges
to and from subgraphs connect to all vertices present in the subgraph. Sugraphs
can also be nested. Some examples of allowed constructs include:
```{r, eval = FALSE}
dag("X -> {Y Z} <- W <-> G")
dag("{X, Y, Z} -> W")
dag("{X -> {Z -> {Y <-> W}}}")
```
These would define the following DAGs:
```mermaid
flowchart LR;
  X(X)-->Y(Y);
  X(X)-->Z(Z);
  X<-.->Z;
  W(W)-->Z;
  W<-.->G(G);
```

```mermaid
flowchart LR;
  X(X)-->W(W);
  Y(Y)-->W;
  Z(Z)-->W;
```

```mermaid
flowchart LR;
  X(X)-->Z(Z);
  X-->Y(Y);
  X-->W(W);
  Z-->Y;
  Z-->W;
  Y<-.->W;
```

## Counterfactual variables and conjunctions

A counterfactual variable is defined by its name, value, and the submodel
it originated from (a set of interventions). For example, $y_x$ is a counterfactual
variable named $Y$ with the value assignment $y$ that originated from a submodel
where the intervention $do(X = x)$ took place.

The function `counterfactual_variable`
and its shorthand alias `cf` can be used to construct counterfactual variables.
This function takes three arguments: `var`, `obs`, and `sub` that correspond
to the variable name, observed value assignment and subscript (the submodel).
For example, $y_x$ is defined as follows:
```{r}
cf(var = "Y", obs = 0, sub = c(X = 0))
```
by default, the value 0 is the "default" or baseline level, and integer values
different from 0 are denoted by primes. For example $y'_x$ is a similar counterfactual
variable to $y_x$, except that it was observed to take the value $y'$ instead of $y$
This can be accomplished by changing the `obs` argument:
```{r}
cf(var = "Y", obs = 1, sub = c(X = 0))
```
Purely observational counterfactual variables (of the original causal model)
can be defined by omitting the `sub` argument.

Conjunctions of multiple counterfactual variables can be constructed using the
function `counterfactual_conjunction` or its shorthand alias `conj`. This
function simply takes an arbitrary number of `"counterfacual_variable"` objects
as its argument. For example, the counterfactual conjunction 
$y \wedge y'_x$ can be defined as follows:
```{r}
v1 <- cf("Y", 0)
v2 <- cf("Y", 1, c("X" = 0))
conj(v1, v2)
```

## Identification

Identifiability of (conditional) counterfactual conjunctions can be determined 
via the function `identifiable`. This function takes the conjunction `gamma` to 
be identified from the set of all interventional distributions $P_*$ of the causal
model represented by the `"dag"` object `g`. An optional conditioning conjunction
`delta` can also be provided. The solution is provided in LaTeX syntax if the
query is identifiable. For instance, we can consider the identifiability
of $P(y_x|x' \wedge z_d \wedge d)$ in the DAG shown below as follows:
```mermaid
flowchart TB;
  X(X)-->W(W);
  W-->Y(Y);
  D(D)-->Z(Z);
  Z-->Y;
  X<-.->Y;
```
```{r}
g1 <- dag("X -> W -> Y <- Z <- D X <-> Y")
v1 <- cf("Y", 0, c(X = 0))
v2 <- cf("X", 1)
v3 <- cf("Z", 0, c(D = 0))
v4 <- cf("D", 0)
c1 <- conj(v1)
c2 <- conj(v2, v3, v4)
identifiable(g = g1, gamma = c1, delta = c2)
```
For more information and examples, please see the package documentation.

## Related packages

 - The [`causaleffect`](https://cran.r-project.org/package=causaleffect) 
   package provides the ID and IDC algorithms for
   the identification of causal effects (among other algorithms).
 - The [`dosearch`](https://cran.r-project.org/package=dosearch) 
   package provides a heuristic search algorithm that uses
   do-calculus to identify causal effects from an arbitrary combination of
   input distributions.
 - The [`dagitty`](https://cran.r-project.org/package=dagitty) provides various
   tools for causal modeling, such as finding adjustment sets and instrumental
   variables.
